---
title: "Beitbridge GPS Exploration"
author: "Sparkgeo"
toc: true
format:
  html:
    theme: zephyr
    html-math-method: katex
    code-tools: true
    self-contained: true
    embed-resources: true
    link-external-icon: true
    link-external-newwindow: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    df-print: kable
execute:
  warning: false
---

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(skimr)
library(lubridate)
library(tidymodels)
library(tidyverse)
```

This notebook is exploring the Beitbridge GPS data with plots and limited statistics. Some of these plots are throw-away.

```{r}
beitbridge_border = read_csv("../data/processed/Beitbridge_Counts_Wait_Time_2018_2022.csv")
```

```{r}
beitbridge_border %>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group = Direction, col = Direction)) + 
  geom_point() +
  scale_color_viridis_d(alpha=0.3,option = "plasma", end = .7) +
  labs(x = "Count", y = "Median Minutes")
```

```{r}
beitbridge_border %>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group= as_factor(StartHour), col = as_factor(StartHour))) + 
  geom_point(alpha = 0.7) +
  scale_color_viridis_d("Hour", option = "plasma", end = .7, alpha =0.2)+
  labs(x = "Count", y = "Median Minutes")
```

# Subsets

## SA - Zimbabwe

```{r}
sa_zimbabwe = beitbridge_border %>% filter(Direction == "SA-Zimbabwe")
```

# Monthly Count

```{r}
 monthly_count_sa = sa_zimbabwe %>% group_by(year=year(StartDate), month=month(StartDate, label=TRUE)) %>% summarize(Month_Total = sum(Count_Events))
```

```{r}
monthly_count_sa %>%
  ggplot(aes(x=month, y=Month_Total, group=as_factor(year), colour=as_factor(year))) +
  geom_line()+
  labs(y = 'Count of Vehicles Processed', x= "Month", caption = "For Vehicles with GPS*", colour ="Year")+
  scale_color_viridis_d(option = "magma", end = .7)
```

We'll start with the South Africa --> Zimbabwe direction. We can tell from the plot above there seems to be higher counts within hour `0`.

We can filter and slice our graph by the hour. Let's look at `0`.

```{r}
sa_zimbabwe %>%
  filter(StartHour == 0 )%>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group= as_factor(StartHour), col = as_factor(StartHour))) + 
  geom_point(alpha = 0.7) +
  scale_color_viridis_d("Hour",option = "plasma", end = .7)+
  labs(x = "Count", y = "Median Minutes")
```


```{r}
sa_zimbabwe %>%
  filter(StartHour == 0 )%>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group = as_factor(year(StartDate)), col = as_factor(year(StartDate)))) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d("Year",option = "plasma", end = .7)+
  labs(x = "Count", y = "Median Minutes")
```

Violin plot zoomed in by start hour 0 and Year.
```{r}
sa_zimbabwe %>%
  filter(StartHour == 0 )%>%
  ggplot(aes(x = as_factor(year(StartDate)), y = Median_Minutes)) +
  geom_violin(alpha = 0.7) +
  scale_color_viridis_d("Year",option = "plasma", end = .7)+
  labs(x = "Year", y = "Median Minutes")
```

```{r}
sa_zimbabwe %>%
  filter(StartHour == 0 )%>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group = as_factor(wday(StartDate, TRUE)), col = as_factor(wday(StartDate, TRUE)))) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d("Day of Week",option = "plasma", end = .7)+
  labs(x = "Count", y = "Median Minutes")
```

```{r}
sa_zimbabwe %>%
  filter(StartHour == 0 )%>%
  ggplot(aes(x = Median_Minutes, y = as_factor(wday(StartDate, TRUE)), group = as_factor(wday(StartDate, TRUE)), col = as_factor(wday(StartDate, TRUE)))) +
  geom_boxplot(alpha = 0.7) +
  scale_color_viridis_d("Day of Week",option = "plasma", end = .7)+
  labs(x = "Count", y = "Median Minutes")
```


## Day of Week

Let's look at day of week across the study period.
```{r}
sa_zimbabwe %>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group= as_factor(wday(StartDate, TRUE)), col = as_factor(wday(StartDate, TRUE)))) + 
  geom_point(alpha = 0.7) +
  scale_color_viridis_d("Day of Week", option = "plasma", end = .7, alpha =0.5)+
  labs(x = "Count", y = "Median Minutes")
```

## Zimbabwe - SA

```{r}
zimbabwe_sa = beitbridge_border %>% filter(Direction == "Zimbabwe-SA")
```


```{r}
zimbabwe_sa %>%
  filter(year(StartDate)=='2019')%>%
  ggplot(aes(x = Count_Events, y = Median_Minutes, group = as_factor(year(StartDate)), col = as_factor(year(StartDate)))) + 
  geom_point(alpha = 0.7) +
   # geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d("Year", option = "plasma", end = .7)+
  labs(x = "Count", y = "Median Minutes")
```

Convert to long format for plotting

```{r}
zim_long = zimbabwe_sa %>%
  select(-Direction)%>%
  pivot_longer(cols = c(Median_Minutes, Bottom_10__Minutes, Top_10__Minutes), values_to = 'minutes')
```

```{r}
zim_long %>%
  filter(StartDate == '2018-01-01' & Count_Events > 0)%>%
  mutate(name = str_replace(name,"_"," "))%>%
   mutate(name = str_replace(name,"__"," "))%>%
  ggplot(aes(x = StartHour, y = minutes, group = name, col = name))+
  geom_point()+
  geom_line() +
  labs(x = "Hour of Day", y = "Minutes", col = "Legend")
```

```{r}
 zimbabwe_sa %>%
  filter(StartDate == '2022-12-30')%>%
  ggplot(aes(x = StartHour, y = Median_Minutes), colour = 'blue')+
  geom_point()+
  geom_line(colour = 'blue')+
  geom_ribbon(aes(ymin = Bottom_10__Minutes, ymax = Top_10__Minutes), fill = 'light blue', alpha = 0.4)+
  labs(x = "Hour", y="Median Minutes")
```

## Daily

```{r}
 zimbabwe_sa %>% group_by(StartDate) %>% summarize(Day_Total = sum(Count_Events))%>%ggplot(aes(x=StartDate, y=Day_Total))+geom_bar(stat = "identity")
```

# Monthly

```{r}
 monthly_count = zimbabwe_sa %>% group_by(year=year(StartDate), month=month(StartDate, label=TRUE)) %>% summarize(Month_Total = sum(Count_Events))
```

```{r}
monthly_count %>%
  ggplot(aes(x=month, y=Month_Total, group=as_factor(year), colour=as_factor(year))) +
  geom_line()+
  labs(y = 'Count of Vehicles Processed', x= "Month", caption = "For Vehicles with GPS*", colour ="Year")+
  scale_color_viridis_d(option = "plasma", end = .7)
```

A clear difference during Covid and lock-downs when aggregated at a monthly scale.